# Linear Models {#linear}

```{r }
#| include: false
library(assertthat)
library(rethinking)
library(brms)
library(dplyr)
library(tidyr)
library(scales)
library(tidybayes)
library(tidybayes.rethinking)
library(posterior)
library(ggplot2)
library(ggdist)
library(modelr)
library(splines)
library(patchwork)
library(paletteer)
```

Some options to facilitate the computations

```{r}
#  For execution on a local, multicore CPU with excess RAM
options(mc.cores = parallel::detectCores())
#  To avoid recompilation of unchanged Stan programs
rstan_options(auto_write = TRUE)
```

The default theme used by `ggplot2`

```{r}
# The default theme used by ggplot2
ggplot2::theme_set(ggdist::theme_ggdist())
ggplot2::theme_update(title = element_text(color = "midnightblue"))
```

Some custom functions used in this following code

```{r}
# Convert widths to probabilities. Useful when plotting.
width2probs <- function(x, add = NULL) {
  stopifnot(is.finite(x), x >= 0.5, x <= 1)
  probs <- (1 - x) / 2
  probs <- c(probs, 1 - probs)
  if (!is.null(add)) probs <- c(probs, add)
  sort(unique(probs))
}
# test the function
stopifnot(all.equal(width2probs(c(0.89, 1), add = 0.5), c(0, 0.055, 0.5, 0.945, 1)))
```





## Why normal distributions are normal

Gaussian distribution

$$
\begin{equation}
P \left(y \mid \mu, \sigma \right) =
\frac{1}{\sqrt{2 \pi} \sigma} \exp{\left[-\frac{1}{2}
 \left(\frac{y-\mu}{\sigma} \right)^2
 \right]}
\end{equation}
$$

gaussian distribution expressed with $precision = \tau$ is $\sigma = \frac{1}{\sqrt{\tau}}$

$$
\begin{equation}
P \left(y \mid \mu, \tau \right) =
\frac{\tau}{\sqrt{2 \pi}} \exp{\left[-\frac{\tau}{2}
 \left(y-\mu \right)^2
 \right]}
\end{equation}
$$

## A language for describing model

$$
\begin{align*}
outcome_i &\sim \mathcal{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta \times predictor_i \\
\beta &\sim \mathcal{Normal}(0, 10) \\
\sigma &\sim \mathcal{HalfCauchy}(0, 1)
\end{align*}
$$

## A Gaussian model of height

### The data

```{r}
data("Howell1")
dataHowel <- Howell1
rm(Howell1)
```

which we can visualize using `skimr`

```{r}
skimr::skim(dataHowel) |>
  mutate(across(.cols = where(is.numeric), .fns = round, digits = 2))
```

select only the adults

```{r}
dataHowel_gte18 <- dataHowel |>
  filter(age >= 18)
```

### The model

$$
\begin{align*}
h_i &\sim \mathcal{N}(\mu, \sigma)\\
\mu &\sim \mathcal{N}(178, 20) \\
\sigma &\sim \mathcal{Uniform}(0, 50)
\end{align*}
$$

We do the prior predictive simulation with the prior $\mu \sim \mathcal{N}(178, 20)$

```{r}
priorHowel_gte18 <- list()
priorHowel_gte18 <- within(priorHowel_gte18, {
  n <- 1e3
  sim1 <- data.frame(id = seq_len(n)) |>
  mutate(mu = rnorm(n = n(), mean = 178, sd = 20),
         sigma = runif(n = n(), min = 0, max = 50),
         height = rnorm(n = n(), mean = mu, sd = sigma))
  })
```

and we do the prior predictive simulation with the prior $\mu \sim \mathcal{N}(178, 100)$

```{r}
priorHowel_gte18 <- within(priorHowel_gte18, {
  n <- 1e4
  sim2 <- data.frame(id=seq_len(1e4)) %>%
    mutate(mu = rnorm(n = n(), mean = 178, sd = 100),
           sigma = runif(n = n(), min = 0, max = 50),
           height = rnorm(n = n(), mean = mu, sd = sigma))
  })
```

and we visualize using `ggplot`. We first create the 2 plots of 
analytical distribution plots

```{r}
plotPrior <- list()
plotPrior$normal <- data.frame(mean = 178, sd = 20) |>
  (\(.){
    ggplot(data=.) +
      geom_function(fun=dnorm, args=list(mean = .$mean, sd = .$sd),
                    color = "olivedrab4", size = 1) +
      scale_x_continuous(limits = c(.$mean - 3 * .$sd, .$mean  + 3 * .$sd),
                     breaks = scales::breaks_width(width = 25)) +
  labs(title = bquote(mu ~ .(sprintf("~ dnorm(%.0f, %.0f)", .$mean, .$sd))), 
       x = expression(mu), y = "density") 
  })()
# plotPrior$normal

plotPrior$uniform <- data.frame(min = 0, max = 50) |>
  (\(.){
    ggplot(data=.) +
      geom_function(fun=dunif, args=list(min=.$min, max=.$max),
                    color="rosybrown2", size=1) +
      scale_x_continuous(limits = c(.$min - 2, .$max  + 2),
                         breaks = scales::breaks_width(width=10)) +
      labs(title = bquote(mu ~ .(sprintf("~ dunif(%.0f, %.0f)", .$min, .$max))), 
           x = expression(sigma), y = "density")
  })()
# plotPrior$uniform
```

we add the actual plot of height to the plots created by McElreath to facilitate
the understanding of priors. McElreath suggests that priors should follows the
scientific knowledge. See the interesting discussion at the end of section 4.3.2
on p. 84.

```{r}
plotPrior <- within(plotPrior, {
  # the density of actual data used for comparisons
  actual <- 
      geom_density(data = dataHowel_gte18, aes(x = height),
                   color = "navy", linewidth = 1, linetype = "dashed")
})
```



```{r}
plotPrior <- within(plotPrior, {
  # NOTE: we use (\(.){})() with |>
  # https://towardsdatascience.com/understanding-the-native-r-pipe-98dea6d8b61b
  sim1 <- priorHowel_gte18$sim1 |>
  (\(.){
    ggplot(data = ., aes(x = height)) +
      geom_density(color = "slateblue1", linewidth = 1) +
      theme(legend.position = c(0.1, 0.8)) +
      labs(
        title = expression(paste("h ~ dnorm(", mu, ",", sigma ,")")),
        subtitle = sprintf("sample size = %d", nrow(.)),
        fill = "quantile")})()
  sim1 <- sim1 + actual  # add the actual data (not in McElreath's curve)
})
# plotPrior$sim1
```

```{r}
plotPrior <- within(plotPrior, {
  sim2 <- priorHowel_gte18$sim2 |>
  (\(.){
    ggplot(data = ., aes(x = height)) +
      geom_density(color = "peru", linewidth = 1) +
      geom_vline(xintercept = 0, linetype = "dotted", color = "navy") +
      theme(legend.position = c(0.1, 0.8)) +
      labs(
        title = expression(paste("h ~ dnorm(", mu, ",", sigma ,")")),
        subtitle = sprintf("sample size = %d", nrow(.)),
        fill = "quantile")
    })()
  sim2 <- sim2 + actual  # add the actual data (not in McElreath's curve)
})
# plotPrior$sim2
```

and putting the plots together with `patchwork` we obtain

```{r}
patchwork::wrap_plots(list(plotPrior$normal, plotPrior$uniform,
                           plotPrior$sim1, plotPrior$sim2)) + 
  plot_annotation(
  title = "Prior Predictive Simulation for the Height Model."
)
```

### Grid approximation of posterior distribution

First create the grid.

```{r}
# create grid of mu and sigma
postHowel_gte18 <- list()
postHowel_gte18 <- within(postHowel_gte18, {
  ngrid <- 200L
  grid <- data.frame(
    mu = seq(from = 140, to = 160, length.out = ngrid),
    sigma = seq(from = 4, to = 9, length.out = ngrid)) |>
    expand(mu, sigma)
  })
```

Then we calculate the likelihood. Since probabilities are percentage this 
causes a numerical issue as multiple multiplications of percentages 
will create very small numbers, so small in fact that they will be miscalculated.

To resolve this problem, we use logarithms.

That is the likelihood function from the model defined in 4.3.2

$$
P(\mu, \sigma \mid h) = 
\prod_{i=1}^n \mathcal{N}(y_i \mid \mu, \sigma) \cdot 
 \mathcal{N}(\mu \mid mean = 0, sd = 10) \cdot 
 \mathcal{U}(\sigma | min = 0, max = 10)
$$

is transformed to log.

> **Important**: Read the end note \# 73 on page 449. All the explanations, including the usage of `max(post$prob)` is explained.

$$
\log{P(\mu, \sigma \mid h)} = 
\sum_{i=1}^n \left[ \log{\mathcal{N}(y_i \mid \mu, \sigma)} +
 \log{\mathcal{N}(\mu \mid mean = 0, sd = 10)} +
 \log{\mathcal{U}(\sigma | min = 0, max = 10)} \right]
$$ 

and to compute the posterior distribution we compute the likelihood which is 
the first element of the addition

$$
\sum_{i=1}^n \log{\mathcal{N}(y_i \mid \mu, \sigma)}
$$ 

as follows

```{r}
# The likelihood on the log scale
postHowel_gte18 <- within(postHowel_gte18, {
  grid <- grid %>%
    mutate(LL = sapply(
      X = seq_len(nrow(.)), 
      FUN = function(i) {
        sum(dnorm(x = dataHowel_gte18$height, 
                  mean = grid$mu[i], 
                  sd = grid$sigma[i], 
                  log = TRUE))
        })
      )
  })
glimpse(postHowel_gte18$grid)
```

then the remaining 2 elements of the summation are the priors

$$
\sum_{i=1}^n \left[
 \log{\mathcal{N}(\mu \mid mean = 0, sd = 10)} +
 \log{\mathcal{U}(\sigma | min = 0, max = 10)} 
 \right]
$$ 

which we add to the likelihood to obtain the posterior distribution on the log scale

```{r}
# add the priors to the likelihood  on the log scales to obtain the
# log of the posterior
postHowel_gte18$grid <- postHowel_gte18$grid |>
  mutate(prob =
           LL + 
           dnorm(x = mu, mean = 178, sd = 20, log = TRUE) +
           dunif(x = sigma, min = 0, max = 50, log = TRUE))
```

and to convert the posterior back to the natural scale we exponentiate.
The usage of `max(the_grid$post)` is explained in endnote 73. 
It is basically used as an approximation to what would be the denominator 
of the likelihood.

$$
\sum_{i=1}^n \left[
 \log{\mathcal{N}(\mu \mid mean = 0, sd = 10)} +
 \log{\mathcal{U}(\sigma | min = 0, max = 10)} 
 \right]
$$

$$
\exp{\left[\log{P(\mu, \sigma \mid h)}\right]} = P(\mu, \sigma \mid h)
$$

```{r}
# convert back to real scale
# attention: see endnote 73 on using max(prob)
postHowel_gte18 <- within(postHowel_gte18, {
  grid$prob <- with(grid, {exp(prob - max(prob))})
})
```

plot the results on a heatmap

```{r}
plotPost <- list()
plotPost$heat <- ggplot(data = postHowel_gte18$grid, aes(x = mu, y = sigma, fill = prob)) +
  geom_raster() +
  scale_x_continuous(limits = c(153, 156)) +
  scale_y_continuous(limits = c(6.5, 9)) +
  scale_fill_paletteer_c("grDevices::Viridis") +
  coord_fixed() +
  labs(title = "The grid's posterior prob.",
       x = expression(mu), y = expression(sigma))
plotPost$heat
```

### Sampling from the grid's posterior

```{r}
postHowel_gte18$draws <- postHowel_gte18$grid |>
  slice_sample(n = 1e4, weight_by = prob, replace = TRUE)
```

and visualizing the density of $\mu$ and $\sigma$ together using `ggExtra`

```{r}
plotPost$marg <- ggplot(data = postHowel_gte18$draws, 
                        mapping = aes(x = mu, y = sigma)) +
  geom_point(color = "mediumorchid", size = 0.8) +
  geom_jitter(color = "mediumorchid", size = 0.8) +
  labs(title = expression("Distribution of" ~ mu ~"and" ~ sigma ~ " using a grid."),
       x = expression(mu), y = expression(sigma))
plotPost$marg <- ggExtra::ggMarginal(plotPost$marg, 
                    xparams = list(colour = "blue", fill = "lightblue", 
                                   linewidth = 1),
                    yparams = list(colour="darkgreen", fill = "lightgreen", 
                                   linewidth = 1))
plotPost$marg
```


### Finding the posterior distribution with `quap` and `brm()`

#### using `rethinking::map`

We now fit the model using `rethinking::quap()`

> See the overthinking box about `list()` vs `alist()` on p. 88 of chapter 4.

The model is

$$
\begin{align*}
h_i &\sim \mathcal{N}(\mu, \sigma)\\
\mu &\sim \mathcal{N}(178, 20) \\
\sigma &\sim \mathcal{Uniform}(0, 50)
\end{align*}
$$

and the fit is

```{r ch04_fit04_01quap}
fit04_01quap <- xfun::cache_rds(
    {rethinking::quap(
      data = dataHowel_gte18,
      flist = alist(
        height ~ dnorm(mu, sigma),
        mu ~ dnorm(178, 28),
        sigma ~ dunif(0, 50)),
      start = list(
        mu  = mean(dataHowel_gte18$height),
        sigma = sd(dataHowel_gte18$height))
      )},
  file = "ch04_fit04_01quap")
```


which gives us the summary

```{r}
precis(fit04_01quap)
```

and the variance covariance matrix is

```{r}
vcov(fit04_01quap) |>
  round(digits = 3)
```

and the correlation matrix

```{r}
cov2cor(vcov(fit04_01quap))
```

#### Using `brms::brm`

This borrows heavily from @kurtz2020b

As mentioned in chapter 8, it is best to use Half-Cauchy distribution for sigma as the tends to work better when using Half Cauchy for sigma when doing a Hamiltonian MCMC with `brm()`.

Therefore the model is

$$
\begin{align*}
h_i &\sim \mathcal{N}(\mu, \sigma)\\
\mu &\sim \mathcal{N}(178, 20) \\
\sigma &\sim \mathcal{HalfCauchy}(0, 1)
\end{align*}
$$

> See the overthinking box about half Cauchy distribution in chapter 8 on p. 260.

This process takes less than a second. It has been save to the rsd file `b04_01.rds`

```{r ch04_fit04_01brm}
tictoc::tic(msg = sprintf("run time of %s, use the cache.", "60 secs."))
fit04_01brm <- xfun::cache_rds({
  brms::brm(data = dataHowel_gte18,
            formula = height ~ 1,
            family = gaussian,
            prior = c(prior(normal(178, 20), class = Intercept),
                      prior(cauchy(0, 1), class = sigma)),
            iter = 2000, warmup = 1000, chains = 4, cores = detectCores(),
            seed = 4)},
  file = "ch04_fit04_01brm")
tictoc::toc()
```


```{r}
plot(fit04_01brm)
```

with the summary

```{r}
summary(fit04_01brm)
```

which can also be done with `tidybayes::summarize_draws` which comes from the
`posterior` package.

```{r}
# normally we only use summarise_draws() but here we change it to match 
# the width of 0.89
tidybayes::summarise_draws(fit04_01brm, mean, median, sd, mad, 
                           ~quantile2(.x, probs = c(0.055, 0.945))) |>
  mutate(across(.cols = where(is.numeric), .fns = round, digits = 2))
```



and to plot the posteriors we need to know the names of the variables

```{r}
tidybayes::get_variables(fit04_01brm)
```

and we spread the data with one column per variable to be able to plot it. 
The `tidybayes` package is particularly useful for this. 
We will use it extensively from now on.

In particular, we can use `tidybayes::spread_draws()` to put variables in 
separate columns or `tidybayes::gather_draws()` to have them in long format.

We can visualize with `ggdist`. it could be done with `tidybayes` but since 
`tidybayes` reexport `ggdist` we use it directly.

```{r ch04_post04_01brm}
post04_01brm <- fit04_01brm |>
    gather_draws(b_Intercept, sigma, ndraws = 500)
```



```{r}
# source: https://cran.r-project.org/web/packages/ggdist/vignettes/slabinterval.html
plot04_01 <- post04_01brm |>
    ggplot(aes(x = .value)) +
    stat_halfeye(aes(fill = stat(level)), .width = c(0.89, 1)) +
    scale_x_continuous(breaks = scales::breaks_extended(n = 7),
                       labels = scales::label_number(accuracy = 0.1)) +
    scale_y_continuous(labels = scales::label_percent()) +
    scale_fill_paletteer_d(palette = "ggsci::teal_material",
                           na.translate = FALSE) +
    theme(legend.position = "none") +
    labs(title = "Model 4.1: Posterior Distribution of Parameters", x = NULL, y = NULL) +
    facet_wrap(. ~ .variable, ncol = 1, scales = "free")
plot04_01
```

### Sampling from a fit

#### Using `quap`

Since `quap` is a quadratic approximation, how do we simulate 2 variables, $\mu$ and $\sigma$?

Simply `quap` gives us the variance covariance. Therefore `quap` can be used 
to simulation the bivariate normal distribution of $\mu$ and $\sigma$

```{r}
vcov(fit04_01quap)
```

from which we can obtain the correlation matrix

```{r}
cov2cor(vcov(fit04_01quap)) |>
  round(digits = 3)
```

so to simulate using `rethinking` we simply use

```{r}
post04_01m <- extract.samples(fit04_01quap, n = 1e4)
```

which gives us a sample of size 10000 of the posterior distribution which can be summarized with the usual `precis()`

```{r}
precis(post04_01m)
```

#### Using `brm`

Using `brm` however we are not given the variance covariance, it is only 
available for the intercept (first-level parameter)

```{r}
vcov(fit04_01brm)
```

So you have to calculate the var-cov matrix by using a sample from the posterior distribution

```{r}
post04_01b <- tidy_draws(fit04_01brm)
# compute the cov
cor(post04_01b[, c("b_Intercept", "sigma")]) |>
  round(digits = 3)
```

> See comment from @kurtz2020b at end of section 4.3.6 to explain that 
McElreath uses `mvnorm()` from `MASS` to simulate using the varcov whereas with 
`brms::tidy_draws()` we do it directly.

Also @kurtz2020b has a nice discussion on how to create summary with histogram.

## Linear predictions

### The linear model strategy

$$
\begin{align*}
h_i &\sim \mathcal{N}(\mu_i, \sigma)\\
\mu_i &= \alpha + \beta (x_i - \bar{x}) \\
\alpha &\sim \mathcal{N}(178, 20) \\
\beta &\sim \mathcal{N}(0,10) \\
\sigma &\sim \mathcal{Uniform}(0, 50)
\end{align*}
$$

#### Probability of the data

$$
h_i \sim \mathcal{N}(\mu_i, \sigma)
$$ 

#### Linear model

$$
\mu_i = \alpha + \beta (x_i - \bar{x})
$$

#### Priors

$$
\begin{align*}
\alpha &\sim \mathcal{N}(178, 20) \\
\beta &\sim \mathcal{N}(0,10) \\
\sigma &\sim \mathcal{Uniform}(0, 50)
\end{align*}
$$

The goal is to **simulate the heights from the model, using only the prior**.

```{r}
priorHeights <- list()
priorHeights <- within(priorHeights, {
  n <- 100L
  set.seed(4)
  sim <- data.frame(
    id = seq_len(n),
    a = rnorm(n = n, mean = 178, sd = 20),
    b = rnorm(n = n, mean = 0, sd = 10)) |>
    expand(nesting(id, a, b), weight = range(dataHowel_gte18$weight)) |>
    mutate(height = a + b * (weight - mean(dataHowel_gte18$weight)))
})
# glimpse(priorHeights$sim)
```

and we plot if

```{r}
ggplot(priorHeights$sim, aes(x = weight, y = height, group = id)) +
  geom_line(alpha = 1/10) +
  geom_hline(yintercept = c(0, 272), linetype = c(2, 1), size = 1/3) +
  coord_cartesian(ylim = c(-100, 400)) +
  labs(title = "b ~ dnorm(0, 10)")
```

##### Adjusting the priors

Since we know that the effect ($\beta$) of the weight on height, 
i.e. the relation between the 2 should be positive and very large value 
unlikely we can use the *log-normal* as a prior on $beta$.

In addition, sigma can also very often be better modeled with the exponential
or HalfCauchy distribution. See section 9.5.3 in the text. 
We will use the exponential distribution for $\sigma$ in this work.

```{r}
priorHeights <- within(priorHeights, {
  lnorm <- ggplot(data.frame(x = c(0, 5)), aes(x)) +
    stat_function(geom = "line", fun = dlnorm, args = list(meanlog = 0, sdlog = 1), 
                color = "slategray", linewidth = 1.5) +
    stat_function(geom = "area", fun = dlnorm, args = list(meanlog = 0, sdlog = 1), 
                fill = "slategray1") +
    scale_y_continuous(labels = scales::label_percent()) +
    labs(title = "log-normal distribution", x = expression(beta), y = "density")
  exp <- ggplot(data.frame(x = c(0, 5)), aes(x)) +
    stat_function(geom = "line", fun = dexp, args = list(rate = 1), 
                color = "seagreen", linewidth = 1.5) +
    stat_function(geom = "area", fun = dexp, args = list(rate = 1), 
                fill = "seagreen1") +
    scale_y_continuous(labels = scales::label_percent()) +
    labs(title = "exponential distribution", x = expression(sigma), y = "density")
})
priorHeights$lnorm + priorHeights$exp
```

$$
\begin{align*}
h_i &\sim \mathcal{N}(\mu_i, \sigma)\\
\mu_i &= \alpha + \beta (x_i - \bar{x}) \\
\alpha &\sim \mathcal{N}(178, 20) \\
\beta &\sim \mathcal{LogNormal}(0,1) \\
\sigma &\sim \mathcal{Exponential}(1)
\end{align*}
$$

### Fitting the posterior distribution

As suggested by the discussion of prior just above, we use a log-normal prior for $\beta$

$$
\begin{align*}
h_i &\sim \mathcal{N}(\mu_i, \sigma)\\
\mu_i &= \alpha + \beta (x_i - \bar{x}) \\
\alpha &\sim \mathcal{N}(178, 20) \\
\beta &\sim \mathcal{LogNormal}(0,1) \\
\sigma &\sim \mathcal{Exponential}(1)
\end{align*}
$$

#### Using `quap`

We add the centralized weight to the data

```{r}
dataHowel_gte18 <- dataHowel_gte18 |>
  mutate(weight_c = as.numeric(scale(weight, center = TRUE, scale = FALSE)))
```

then get the fit using `rethinking::quap`

> Giving start values to `quap` seem to help it significantly and avoiding error, 
at least when using b \~ dlnorm(0, 1).


```{r ch04_fit04_03quap}
tictoc::tic()
fit04_03quap <- xfun::cache_rds(
    {rethinking::quap(
      data = dataHowel_gte18,
      flist = alist(
      height ~ dnorm(mu, sigma),
      mu <- a + b * weight_c,
      a ~ dnorm(178, 20),
      b ~ dlnorm(0, 1),
      sigma ~ dunif(0, 50)),
    start = list(
      a  = mean(dataHowel_gte18$height),
      sigma = sd(dataHowel_gte18$height))
    )},
  file = "ch04_fit04_03quap")
tictoc::toc()
```

```{r}
precis(fit04_03quap)
```

#### Using `brm`

Again, we use the exponential distribution as a prior of sigma to facilitate the iterations with `brm`. There are 2 equivalent ways to run this model. One uses the log-normal distribution of $\beta$, the other one uses the log transform of $\beta$ with the normal distribution. The two models are mathematically equivalent

#### Using lognormal distribution

$$
\begin{align*}
h_i &\sim \mathcal{N}(\mu_i, \sigma)\\
\mu_i &= \alpha + \beta (x_i - \bar{x}) \\
\alpha &\sim \mathcal{N}(178, 20) \\
\beta &\sim \mathcal{LogNormal}(0,1) \\
\sigma &\sim \mathcal{Exponential}(1)
\end{align*}
$$

When using lognormal for a parameter of class b, you should specify lb and ub (lower bound and upper bound) to avoid error message and accelerate the computations with `brm`.

```{r ch04_fit04_03brm}
tictoc::tic(msg = sprintf("run time of %s, use the cache.", "70 secs."))
fit04_03brm <- xfun::cache_rds({
  brms::brm(
    data = dataHowel_gte18,
    family = gaussian,
    formula = height ~ 1 + weight_c,
    prior = c(
      prior(normal(178, 20), class = Intercept),
      prior(lognormal(0, 1), class = b, lb = 0, ub = 3),
      prior(exponential(1), class = sigma)),
    iter = 2000, warmup = 1000, cores = detectCores(), chains = detectCores(),
    seed = 4)},
  file = "ch04_fit04_03brm")
tictoc::toc()
```

```{r}
summarize_draws(fit04_03brm) |>
  mutate(across(.cols= where(is.numeric), .fns = round, digits = 1))
```

### Using the log tranformation

$$
\begin{align*}
h_i &\sim \mathcal{N}(\mu_i, \sigma)\\
\mu_i &= \alpha + \exp{(log\_b)} (x_i - \bar{x}) \\
\alpha &\sim \mathcal{N}(178, 20) \\
log\_b &\sim \mathcal{N}(0,1) \\
\sigma &\sim \mathcal{Exponential}(1)
\end{align*}
$$

```{r}
glimpse(dataHowel_gte18)
```


```{r ch04_fit04_03brm_b}
tictoc::tic(msg = sprintf("run time of %s, use the cache.", "60 secs."))
fit04_03brm_b <- xfun::cache_rds({
  brms::brm(
  data = dataHowel_gte18,
  family = gaussian,
  formula = bf(height ~ a + exp(lb) * weight_c,
               a ~ 1, lb ~ 1, nl = TRUE),
  prior = c(
    prior(normal(178, 20), class = b, nlpar = a),
    prior(normal(0, 1), class = b, nlpar = lb),
    prior(exponential(1), class = sigma)),
  iter = 2000, warmup = 1000, chains = 4, cores = detectCores(), seed = 4)},
  file = "ch04_fit04_03brm_b")
tictoc::toc()
```

```{r}
summarize_draws(fit04_03brm_b) |>
  mutate(across(.cols = where(is.numeric), .fns = round, digits = 2))
```

### Interpreting the posterior distribution

#### Tables of marginal distributions

Using `rethinking` **Important**, the parameters are correlated here, 
to avoid this one must do **centering** of variables. 
The following uses **centered** variables.

```{r}
precis(fit04_03quap, corr = TRUE)
```

```{r}
round(vcov(fit04_03quap), 3)
```

Using `brm`

Note: `lp__` stands for *unnormalized log posterior density*.

```{r}
# normally we only use summarise_draws() but here we change it to match 
# the width of 0.89
tidybayes::summarise_draws(fit04_03brm, mean, median, sd, mad, 
                           ~quantile2(.x, probs = c(0.055, 0.945)),
                           default_convergence_measures(),
                           default_mcse_measures()) |>
  mutate(across(.cols = where(is.numeric), .fns = round, digits = 2))
```

we get the varcov matrix as follows

```{r}
tidy_draws(fit04_03brm) |>
  select(!matches("^[.]|__$")) |>
  cov() |>
  round(digits = 3)
```

and the correlation matrix

```{r}
tidy_draws(fit04_03brm) |>
  select(!matches("^[.]|__$")) |>
  cor() |>
  round(digits = 3)
```

#### Plotting posterior inference against data

With `brms` we use the `ggmcmc` package to illustrate the results from the markov chain

```{r}
tidybayes::get_variables(fit04_03brm)
```

```{r ch04_post04_03brm}
post04_03brm <- list()
post04_03brm <- within(post04_03brm, {
  long <- fit04_03brm |>
    tidybayes::gather_draws(b_Intercept, b_weight_c, sigma)
})
```

with the histogram

```{r}
plot04_03brm <- list()
plot04_03brm$hist <- ggplot(post04_03brm$long, aes(x = .value)) +
  geom_histogram(aes(fill = .variable)) +
  scale_fill_paletteer_d(palette = "futurevisions::atomic_orange") +
  theme(legend.position = "none") +
  facet_wrap(. ~ .variable, ncol = 1, scales = "free")
plot04_03brm$hist
```

and density plots by chains

```{r}
plot04_03brm$dens <- ggplot(post04_03brm$long, 
                            aes(x = .value, color = as.factor(.chain))) +
  geom_density() +
  scale_color_paletteer_d(palette = "futurevisions::atomic_clock") +
  labs(x = NULL, color = "chain") +
  facet_wrap(. ~ .variable, ncol = 1, scales = "free")
plot04_03brm$dens
```

and the paired plots with `ggally`

```{r}
post04_03brm <- within(post04_03brm, {
  wide <- fit04_03brm |>
    tidybayes::spread_draws(b_Intercept, b_weight_c, sigma)
})
```

```{r}
plot04_03brm <- within(plot04_03brm, {
  fun_diag <- function(data, mapping, ...){
    ggplot(data = data, mapping = mapping) +
      geom_density(linewidth = 1)
  }
  fun_lower <- function(data, mapping) {
    ggplot(data = data, mapping = mapping) +
      stat_density2d(linewidth = 1/3)
  }
  pairs <- GGally::ggpairs(
    data = post04_03brm$wide, 
    mapping = aes(color = as.factor(.chain)),
    columns = c("b_Intercept", "b_weight_c", "sigma"),
    diag = list(continuous = fun_diag),
    lower = list(continuous = fun_lower)) +
    scale_color_paletteer_d(palette = "futurevisions::atomic_clock", direction = -1) +
    # scale_fill_paletteer_d(palette = "futurevisions::atomic_clock") +
    labs(title = "Parameters Comparisons by Chain")
})
plot04_03brm$pairs
```



and the correlation matrix

```{r}
plot04_03brm$corr <- post04_03brm$wide |>
  select(b_Intercept, b_weight_c, sigma) |>
  GGally::ggcorr(color = "darkgreen",
                 nbreaks = 10, label = TRUE, label_round = 2,
                 label_color = "midnightblue") +
  scale_fill_paletteer_d(palette = "futurevisions::venus") +
  theme(legend.position = "none") +
  labs(title = "Correlations between parameters")
plot04_03brm$corr
```

and for added extra, the trace plot

```{r}
plot04_03brm$trace <- post04_03brm$long |>
  ggplot(aes(x = .iteration, y = .value, color = as.factor(.chain))) +
  geom_line() +
  scale_color_paletteer_d(palette = "futurevisions::atomic_clock", direction = 1) +
  theme(legend.position = "none") +
  labs(x = NULL, y = NULL) +
  facet_wrap(. ~ .variable, ncol = 1, scales = "free")
plot04_03brm$trace
```

## Curves from lines

### Polynomial regression

```{r}
data("Howell1")
dataHowel <- Howell1 |>
  # use as.vector() to keep the attribute
  mutate(weight_s = scale(as.vector(weight)),
         weight_s2 = weight_s ^ 2)
rm(Howell1)
```

```{r}
plotHowel <- list()
plotHowel <- within(plotHowel, {
  colr <- unclass(paletteer::paletteer_d("futurevisions::titan"))
  basic <- dataHowel |>
    (\(.) {
      ggplot(., aes(x = weight_s, y = height, color = age)) +
        scale_x_continuous(breaks = scales::breaks_extended(n=7),
                     labels = function(x) {
                       x <- x * sd(.$weight) + mean(.$weight)
                       label_number(accuracy = 1)(x)
                     }) +
        scale_color_gradientn(colors = colr) +
        geom_point(shape = 20, size = 2, alpha = 2/3) +
        theme(legend.position = c(0.1, 0.8)) +
        labs(title = "Census data for the Dobe area !Kung San",
       subtitle = sprintf("%d individuals", nrow(.)))
      })()
})
plotHowel$basic
```

and the model used is

$$
\begin{align*}
h_i &\sim \mathcal{N}(\mu_i, \sigma)\\
\mu_i &= \alpha + \beta_1 \cdot weight\_s_i + \beta_2 \cdot weight\_s^2_i \\
\alpha &\sim \mathcal{N}(178, 20) \\
\beta_1 &\sim \mathcal{LogNormal}(0,1) \\
\beta_2 &\sim \mathcal{N}(0,1) \\
\sigma &\sim \mathcal{Exponential}(1)
\end{align*}
$$

https://discourse.mc-stan.org/t/error-with-gamma-prior/16420

::: callout-warning
The following code gives a warning about setting lower boundaries. It started to show with `R 4.2`. Paul Buerkner advises to ignore it. See [advice](https://discourse.mc-stan.org/t/error-with-gamma-prior/16420)
:::

```{r ch04_fit04_05brm}
tictoc::tic(msg = sprintf("run time of %s, use the cache.", "60 secs."))
fit04_05brm <- xfun::cache_rds({
  brm(data = dataHowel,
      family = gaussian,
      height ~ 1 + weight_s + weight_s2,
      prior = c(prior(normal(178, 20), class = Intercept),
                prior(lognormal(0, 1), class = b, coef = "weight_s"),
                prior(normal(0, 1), class = b, coef = "weight_s2"),
                prior(exponential(1), class = sigma)),
      iter = 4000, warmup = 2000, chains = 4, cores = detectCores(),
      seed = 4)},
  file = "ch04_fit04_05brm")
tictoc::toc()
```

```{r}
summarize_draws(fit04_05brm) |>
  filter(!grepl(pattern = "__$", x = variable)) |>
  mutate(across(.cols = where(is.numeric), .fns = round, digits = 2))
```

and to obtain a simplified dataframe we use

```{r}
brms::fixef(fit04_05brm) |>
  round(digits = 2)
```

```{r}
tidybayes::get_variables(fit04_05brm)
```

```{r ch04_post04_05brm}
post04_05brm <- xfun::cache_rds({
  fit04_05brm |> 
    tidybayes::gather_draws(b_Intercept, b_weight_s, b_weight_s, sigma)},
  file = "ch04_post04_05brm")
```




```{r}
plot04_05brm <- list()
plot04_05brm$dens <- 
  post04_05brm |>
  ggplot(aes(x = .value, color = as.factor(.chain))) +
  geom_density() +
  scale_color_paletteer_d(palette = "futurevisions::mars", direction = -1) +
  theme(legend.position = "none") +
  labs(x = NULL, color = "chain") +
  facet_wrap(. ~ .variable, ncol = 1, scales = "free")
```

and

```{r}
plot04_05brm$trace <- 
  post04_05brm |>
  ggplot(aes(x = .iteration, y = .value, color = as.factor(.chain))) +
  geom_line() +
  scale_color_paletteer_d(palette = "futurevisions::mars", direction = -1) +
  theme(legend.position = "none") +
  labs(x = NULL, y = NULL) +
  facet_wrap(. ~ .variable, ncol = 1, scales = "free")
```

```{r}
plot04_05brm$dens + plot04_05brm$trace
```

And we look at the fitted and predicted values to understand and interpret the result.

What is the difference between *fitted* and *predict*? *fitted* A nice explanation is given by [Greg Snow](https://stackoverflow.com/questions/12201439/is-there-a-difference-between-the-r-functions-fitted-and-predict)

> The `fitted` function returns the y-hat values associated with the data used to fit the model. The `predict` function returns predictions for a new set of predictor variables. If you don't specify a new set of predictor variables then it will use the original data by default giving the same results as `fitted` for some models (especially the linear ones), but if you want to predict for a new set of values then you need `predict`. The `predict` function often also has options for which type of prediction to return, the linear predictor, the prediction transformed to the response scale, the most likely category, the contribution of each term in the model, etc.

Therefore, if we give the same data to `fitted` or `predict` will will obtain sensibly the same results, the difference being caused by the random seed. However, in Bayesian stats, `fitted` will only provide $\mu_i$ and its variation whereas `predict` will give $h_i$ which is $h_i \sim \mathcal{N}(\mu_i, \sigma)$

We can see it clearly here as `fitd_quad` gives ans estimate about the same as for `predict` since they both report the same `\mu_i`, but `predict` has a wider interval since it uses $\sigma$

```{r}
pred04_05brm <- list()
pred04_05brm <- data.frame(
  weight_s = seq_range(dataHowel$weight_s, n = 30L)) |>
  mutate(weight_s2 = weight_s^2) |>
  add_predicted_draws(fit04_05brm, ndraws = 500) |>
  median_qi()

lpred04_05brm <- data.frame(
  weight_s = seq_range(dataHowel$weight_s, n = 30L)) |>
  mutate(weight_s2 = weight_s^2) |>
  add_linpred_draws(fit04_05brm, ndraws = 500) |>
  median_qi()
```

and we can now create the plot.

```{r}
plot04_05brm$model <- plotHowel$basic +
  geom_ribbon(data = pred04_05brm,
              aes(x = weight_s, ymin = .lower, ymax = .upper),
              inherit.aes = FALSE, fill = "lightcyan", alpha = 1) +
  geom_smooth(data = lpred04_05brm,
              aes(x=weight_s, y = .linpred, ymin = .lower, ymax = .upper),
              inherit.aes = FALSE, stat = "identity",
              fill = "lightcyan3", color = "royalblue", alpha = 1, size = 1/2) +
  geom_point(shape = 20, size = 2, alpha = 2/3)
plot04_05brm$model
```

### Splines

```{r}
data("cherry_blossoms")
dataCherry <- cherry_blossoms
rm(cherry_blossoms)
dataCherry |> skimr::skim()
# data without NA
dataCherry_nona <- dataCherry |>
  drop_na(doy)
```

#### Knots, degree and basis functions

The knots used here are based on quantiles, other ways are possible,

```{r}
knots <- quantile(dataCherry_nona$year, probs = seq(from = 0, to = 1, length.out = 15))
knots
```

```{r}
colr <- unclass(paletteer::paletteer_d("futurevisions::cancri"))
ggplot(dataCherry_nona, aes(x = year, y = doy, color = temp)) +
  geom_vline(xintercept = knots, color = "slateblue", alpha = 1/2) +
  geom_point(shape = 20, size = 2, alpha = 2/3) +
  scale_x_continuous(breaks = knots, labels = scales::label_number(big.mark = "")) +
  scale_color_gradientn(colors = colr) +
  theme(legend.position = c(0.05, 0.8),
        axis.text.x = element_text(size = rel(0.9))) +
  labs(title = "Cherry Blossom in Japan",
       subtitle = sprintf("%d observations with %d knots", nrow(dataCherry_nona), length(knots)))
```

the code `knots[-c(1, nknots)]` is required because `bs` places knots at 
the boundaries by default, so we have to remove them.

```{r}
B <- splines::bs(x = dataCherry_nona$year, 
                 knots = knots[-c(1, length(knots))], 
                 degree = 3, intercept = TRUE)
# str(B)
```

and we plot the basis functions

```{r}
# this data.frame will be reused below with the posteriors
df_bias <- B |>
  as.data.frame() %>%
  setNames(sprintf("B%02d", seq_len(ncol(.)))) |>
  mutate(year = dataCherry_nona$year) |>
  pivot_longer(cols = -year, names_to = "bias_func", values_to = "bias")
# str(df_bias)

clrs <- paletteer::paletteer_c("pals::jet", n = length(unique(df_bias$bias_func)))
ggplot(df_bias, aes(x = year, y = bias, color = bias_func)) +
  geom_vline(xintercept = knots, color = "grey60", linetype = "longdash", alpha = 1/2) +
  geom_line() +
  scale_x_continuous(breaks = knots, labels = scales::label_number(big.mark = "")) +
  scale_color_manual(values = clrs) +
  theme(legend.position = "none") +
  labs("The bias functions")
```

#### Model and fit

$$
\begin{align*}
doy_i &\sim \mathcal{N}(\mu_i, \sigma) \\
\,u_i &= \alpha + \sum_{k=1}^Kw_kB_{k, i} \\
\alpha &\sim \mathcal{N}(100, 10) \\
w_j &\sim \mathcal{N}(0, 10) \\
\sigma &\sim \mathcal{Exp}(1)
\end{align*}
$$

We first append the matrix to the data in one column. See @kurtz2020b on this data structure.

```{r}
dataCherry_nonaB <- dataCherry_nona |>
  mutate(B = B)
# the last column is a matrix column, with same nb of rows as the other
# columns but with a column including 17 subcolumns (!)
```

and the fit

```{r ch04_fit04_08brm}
tictoc::tic(msg = sprintf("run time of %s, use the cache.", "70 secs."))
fit04_08brm <- xfun::cache_rds({
  brm(data = dataCherry_nonaB,
      family = gaussian,
      doy ~ 1 + B,
      prior = c(prior(normal(100, 10), class = Intercept),
                prior(normal(0, 10), class = b),
                prior(exponential(1), class = sigma)),
      cores = detectCores(), seed = 4)},
  file = "ch04_fit04_08brm")
tictoc::toc()
```

```{r}
summarize_draws(fit04_08brm) |>
  filter(!grepl(pattern = "__$", x = variable)) |>
  mutate(across(.cols = where(is.numeric), .fns = round, digits = 1))
```

#### Plot

```{r}
get_variables(fit04_08brm)
```

```{r}
# Source: https://github.com/mjskay/tisdybayes/issues/38
post04_08brm <- tidybayes::gather_draws(fit04_08brm, !!sym("^b_B.+"), regex = TRUE) |>
  mutate(.variable = as.integer(sub("^b_B", replacement = "", x = .variable)),
         .variable = sprintf("B%02d", .variable)) |>
  rename("bias_func" = .variable) |>
  group_by(bias_func) |>
  summarise(weight = mean(.value)) |>
  full_join(y = df_bias, by = "bias_func")
# glimpse(df)

plot04_08brm <- list()
plot04_08brm <- within(plot04_08brm, {
  clrs <- paletteer::paletteer_c("pals::jet", n = length(unique(post04_08brm$bias_func)))
  bias <- ggplot(post04_08brm, aes(x = year, y = bias * weight, color = bias_func)) +
    geom_vline(xintercept = knots, color = "grey60", linetype = "longdash", alpha = 1/2) +
    geom_line(size = 1) +
    scale_x_continuous(breaks = knots, labels = scales::label_number(big.mark = "")) +
    scale_color_manual(values = clrs) +
    theme(legend.position = "none") +
    labs(title = "fitted bias functions")
})
plot04_08brm$bias
```
the fitted values

```{r}
lpred04_08brm <- dataCherry_nonaB |> 
    add_linpred_draws(fit04_08brm) |>
    ungroup() |>
    select(-B) |>
    group_by(year) |>
    mean_qi(.linpred)
```

and the plot


```{r}
plot04_08brm <- within(plot04_08brm, {
  clrs <- unclass(paletteer::paletteer_d("futurevisions::cancri"))
  p <- ggplot(lpred04_08brm, aes(x = year, y = .linpred)) +
    geom_vline(xintercept = knots[-c(1, length(knots))], color = "slateblue", alpha = 1/2) +
    geom_point(dataCherry_nonaB, mapping = aes(x = year, y = doy, color = temp),
             inherit.aes = FALSE) +
    geom_lineribbon(aes(x = year, y = .linpred, ymin = .lower, ymax = .upper),
                  color = "blueviolet", fill = "cornflowerblue", alpha = 1/2) +
    scale_x_continuous(breaks = knots, labels = scales::label_number(big.mark = "")) +
    scale_color_gradientn(colors = clrs) +
    theme(legend.position = "none") +
    labs(title = "Figure 4.12",x = "year", y = "doy")
})
plot04_08brm$p
```



### Smooth functions for a smooth world

See @kurtz2020b for much more details on this topic.

## Summary

This was an important chapter. Most of the plots and basic coding tools are 
exemplified here. It is an important reference chapter. 
The `brms` package will be exclusively used from now on.
